\bf A Comparison of Parametric and Permutation Tests for Regression Analysis of Randomized Experiments

### 2017-09-11 ms/ms.tex
Edits to text, add refs

To carry out the standard parametric hypothesis test for a linear model, the following assumptions are needed (\cite{freedman_statistical_2005}):
\todo{CHECK}
\begin{enumerate}
\item \textbf{Linearity:} The data $Y$ are related to $X$ and $Z$ linearly.
\item \textbf{Constant slopes:} Stratum membership only affects the intercept $\alpha_j$, not the slopes $\beta$ and $\gamma$.
\item \textbf{IID Errors:} The $\varepsilon_{ij}$ are independent and identically distributed with mean $0$ and common variance $\sigma^2$.
\item \textbf{Independence:} If $X$ is random, $\varepsilon$ is statistically independent of $X$.
\item \textbf{Normality:} The errors are normally distributed.
\end{enumerate}

\todo{are the estimator, standard error, variance clear here?}
The coefficients are estimated using least squares (or equivalently, by maximizing the likelihood).
The estimated coefficient $\hat{\gamma}$ is the estimated average treatment effect. 
(Note, only Assumptions (1) and (2) are needed to guarantee the existence of a solution.
The solution will be unique as long as there is no linear relationship between $X$, $Z$, and stratum membership.)
This procedure also yields an estimate $\hat{\sigma}_{\hat{\gamma}}^2$ of the variance of $\hat{\gamma}$.
Under the null hypothesis, the test statistic 
$$ T = \frac{\hat{\gamma}}{\sqrt{ \hat{\sigma}_{\hat{\gamma}}^2}}$$
follows the Student $t$ distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
The $p$-value for this hypothesis test is the probability, assuming the null hypothesis of zero coefficient is true, that a value drawn from the $t$ distribution is larger in magnitude than $T$.
This test is equivalent to the $F$-test and the likelihood ratio test assuming the Gaussian model,
and \todo{ cite this? is therefore locally most powerful ????}.

However, in the context of a randomized experiment, these assumptions will not be satisfied.
The estimated treatment effect $\hat{\gamma}$ and nominal standard errors $\hat{\sigma}_{\hat{\gamma}}$ can be severely biased (\todo{cite freedman 2008, lin 2013}).
\todo{Lin 2013} shows that running a regression with a full set of treatment and covariate interaction terms cannot hurt asymptotic precision, and using the Huber-White sandwich standard errors can yield asymptotically valid confidence intervals.
In small samples, the bias may still be substantial.
\todo{cite Sekhon post-stratification} show that post-stratification, estimating treatment effects within groups of similar individuals defined after data are collected, can ameliorate this bias.
This is equivalent to estimating a fully saturated linear model with interaction terms for treatment and stratum membership.
The linear ANCOVA model does not account for variation in the treatment effect across strata in this way;
if the difference in potential outcomes is heterogeneous across strata, then the coefficient $\gamma$ may be attenuated towards 0.

\subsection{Stratified permutation test}\label{subsec:strat_perm_test}
In randomization inference, the only random quantities are those involving $Z$, the vector of treatment assignments.
The exact inference is derived from the randomization distribution of statistics involving $Z$.
In particular, the potential outcomes and covariates $(Y_{ij}(1), Y_{ij}(0), X_{ij})$ are fixed in the finite population under study, while the observed response $Y_{ij}$ changes with $Z_{ij}$.
\todo{cite rosenberg 2002}

@@

Under the null hypothesis, which treatment an individual received amounts to an arbitrary label.
Once we observe one response under a particular treatment, we can impute the other potential outcome; namely, it would have been the same.
Thus, the observed $Y_{ij}$ form a sufficient statistic for the data. \todo{cite P\&S}
This null hypothesis is stronger than the null hypothesis for the parametric ANCOVA, which only asserts that differences in potential outcomes are zero \textit{on average}.

@@

The treatment effect for individual $(i, j)$ is $\gamma e^{v_{ij}}$ and we would like to estimate the average treatment effect in the sample.

Assume that there are $n_j=16$ individuals per stratum and treatment assignment is balanced, i.e. 8 people receive each treatment at each stratum.
After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerate $Z$ and the potential outcomes $Y(Z)$ 10,000 times.
We repeat this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.

@@

>>>>

To carry out the standard parametric hypothesis test for a linear model, the following assumptions are needed (\cite{freedman_statistical_2005}):
\begin{enumerate}
\item \textbf{Linearity:} The data $Y$ are related to $X$ and $Z$ linearly.
\item \textbf{Constant slopes:} Stratum membership only affects the intercept $\alpha_j$, not the slopes $\beta$ and $\gamma$.
\item \textbf{IID Errors:} The $\varepsilon_{ij}$ are independent and identically distributed with mean $0$ and common variance $\sigma^2$.
\item \textbf{Independence:} If $X$ is random, $\varepsilon$ is statistically independent of $X$.
\item \textbf{Normality:} The errors are normally distributed.
\end{enumerate}

The coefficients are estimated using least squares (or equivalently, by maximizing the likelihood).
The estimated coefficient $\hat{\gamma}$ is the estimated average treatment effect. 
This procedure also yields an estimate $\hat{\sigma}_{\hat{\gamma}}^2$ of the variance of $\hat{\gamma}$.
Under the null hypothesis, the test statistic 
$$ T = \frac{\hat{\gamma}}{\sqrt{ \hat{\sigma}_{\hat{\gamma}}^2}}$$
follows the Student $t$ distribution with degrees of freedom equal to the number of observations minus the number of parameters estimated (in this case, $N - J - 2$).
The $p$-value for this hypothesis test is the probability, assuming the null hypothesis of zero coefficient is true, that a value drawn from the $t$ distribution is larger in magnitude than $T$.
This test is equivalent to the $F$-test.
When the model assumptions are true, it is uniformly most powerful.

However, in the context of a randomized experiment, these assumptions will not be satisfied.
The estimated treatment effect $\hat{\gamma}$ and nominal standard errors $\hat{\sigma}_{\hat{\gamma}}$ can be severely biased (\cite{freedman_regression_2008, lin_agnostic_2013}).
\cite{lin_agnostic_2013} shows that running a regression with a full set of treatment and covariate interaction terms cannot hurt asymptotic precision, and using the Huber-White sandwich standard errors can yield asymptotically valid confidence intervals.
In small samples, the bias may still be substantial.
\cite{miratrix_adjusting_2013} show that post-stratification, estimating treatment effects within groups of similar individuals defined after data are collected, can ameliorate this bias.
This is equivalent to estimating a fully saturated linear model with interaction terms for treatment and stratum membership.
The linear ANCOVA model does not account for variation in the treatment effect across strata in this way;
if the difference in potential outcomes is heterogeneous across strata, then the coefficient $\gamma$ may be attenuated towards 0.

\subsection{Stratified permutation test}\label{subsec:strat_perm_test}
In randomization inference, the only random quantities are those involving $Z$, the vector of treatment assignments.
The exact inference is derived from the randomization distribution of statistics involving $Z$.
In particular, the potential outcomes and covariates $(Y_{ij}(1), Y_{ij}(0), X_{ij})$ are fixed in the finite population under study, while the observed response $Y_{ij}$ changes with $Z_{ij}$.

@@

Under the null hypothesis, which treatment an individual received amounts to an arbitrary label.
Once we observe one response under a particular treatment, we can impute the other potential outcome; namely, it would have been the same  (\cite{rosenbaum_covariance_2002}).
Thus, the observed $Y_{ij}$ form a sufficient statistic for the data (\cite{pesarin_permutation_2010}).
This null hypothesis is stronger than the null hypothesis for the parametric ANCOVA, which only asserts that differences in potential outcomes are zero \textit{on average}.

@@

We included $n_j=16$ individuals per stratum and treatment assignment was balanced, i.e. 8 people received each treatment in each stratum.
After sampling $(v_{ij}, \varepsilon_{ij}, \delta_{ij})$, we regenerated $Z$ and the potential outcomes $Y(Z)$ 10,000 times.
We repeated this procedure for each distribution of latent variables $v$ and of the errors $\varepsilon$ and $\delta$.

The treatment effect for individual $(i, j)$ was $\gamma e^{v_{ij}}$.
We tested whether the treatment had a nonzero effect in the sample using each of the tests described.

@@

Many outcomes are measured on a discrete, ordinal scale but the same regression-based methods are used to test for a nonzero treatment effect.
We repeated the procedure above, instead discretizing the covariate $X$ and potential outcomes $Y(Z)$ by taking their floor.
We used the same distributions for the latent variables $v$.
We drew $\varepsilon$ and $\delta$ from independent standard normal distributions.
